---
title: "Results WNG (final)"
author: "Piotr Szefer"
date: "July 5, 2018"
output: pdf_document
---

```{r global.options, include = TRUE, echo=FALSE}
knitr::opts_chunk$set(
    fig.align   = 'center',
    fig.height  = 3,
    echo        = FALSE,
    message     = FALSE,
    warning     = FALSE)
```

# Community composition

## RDA

```{r, fig.width=8,fig.height=4}
source("C:/Users/Piotr Szefer/Desktop/Work/garden experiment/code/rda_plots.R")

# RDA plots and tests
library(cowplot)
plot_grid(pAll, pTree)
# pAll
# pTree
```
Fig. Proportional data, full biomass block covariate, standarization by samples.
```{r}
library(knitr)
kable(p1axis, caption = "Permutation test results for the whole community")
kable(p2axis, caption = "Permutation test results for all tree species")
```

```{r, echo = FALSE, include = FALSE, message=FALSE, warning=FALSE}
#source("C:\\Users\\Piotr Szefer\\Desktop\\Work\\garden experiment\\code\\Data_Processing_Script_v2.R")

source("C:\\Users\\Piotr Szefer\\Desktop\\Work\\garden experiment\\code\\Data_Processing_Script_v3.R")
```

# Community descriptors

## For the whole community and for the tree community
```{r, fig.height = 6, fig.width=6}

panelData$cols <- rep("grey40", dim(panelData)[1])
# What is significant and what is marginally significant.

# Get rid of uninformative Simpsons D
panelData <- panelData[!(panelData$TYPE %in% c("Simpson's D","Simpson's D (Woody plants)")),]

sig <- "red"
msig <- "orange"

# Whole community
# summary(l1W) # Biomass Weevil 125
panelData[(panelData$TYPE == "LN(Biomass)" & panelData$TREAT == "H2"), ]$cols <- sig
panelData[(panelData$TYPE == "LN(Biomass)" & panelData$TREAT == "P"), ]$cols <- msig

# summary(l2W) # Shannon Insecticide and marginally weevil 25
panelData[(panelData$TYPE == "Shannon's H" & panelData$TREAT == "I"), ]$cols <- sig
panelData[(panelData$TYPE == "Shannon's H" & panelData$TREAT == "H1"), ]$cols <- msig
# summary(l3W) # Species number: Insecticide
panelData[(panelData$TYPE == "Richness" & panelData$TREAT == "I"), ]$cols <- msig
# summary(l4W) # Simpsons: Insecticide
# panelData[(panelData$TYPE == "Simpson's D" & panelData$TREAT == "I"), ]$cols <- sig
# summary(l5W) # Evenness: marginally insecticide and weevil 25
panelData[(panelData$TYPE == "Evenness" & panelData$TREAT == "I"), ]$cols <- msig
panelData[(panelData$TYPE == "Evenness" & panelData$TREAT == "H1"), ]$cols <- msig

# "LN(Biomass) (Woody plants)"
# "Shannon's H (Woody plants)"
# "Richness (Woody plants)"
# "Simpson's D (Woody plants)"
# "Evenness (Woody plants)"
# "Number of stems (Woody plants)"

# Trees
# summary(l1Wt) # Biomass: Weevil 125
panelData[(panelData$TYPE == "LN(Biomass) (Woody plants)" & panelData$TREAT == "H2"), ]$cols <- sig

# summary(l2Wt) # Shannon: marginally Insecticide
panelData[(panelData$TYPE == "Shannon's H (Woody plants)" & panelData$TREAT == "I"), ]$cols <- msig

# summary(l3Wt) # Species number: marginally Insecticide and marginially FUNGICIDE
# panelData[(panelData$TYPE == "Richness (Woody plants)" & panelData$TREAT == "I"), ]$cols <- msig
panelData[(panelData$TYPE == "Richness (Woody plants)" & panelData$TREAT == "F"), ]$cols <- msig

# summary(l4Wt) # Simpsons: marginally Insecticide
# panelData[(panelData$TYPE == "Simpson's D (Woody plants)" & panelData$TREAT == "I"), ]$cols <- msig
# summary(l5Wt) # Evenness: marginally predator
panelData[(panelData$TYPE == "Evenness (Woody plants)" & panelData$TREAT == "P"), ]$cols <- msig

# summary(l6Wt) # Stems: fungicide, predator and weevil 125
panelData[(panelData$TYPE == "No. of stems (Woody plants)" & panelData$TREAT == "F"), ]$cols <- sig
panelData[(panelData$TYPE == "No. of stems (Woody plants)" & panelData$TREAT == "P"), ]$cols <- sig
panelData[(panelData$TYPE == "No. of stems (Woody plants)" & panelData$TREAT == "H2"), ]$cols <- sig

panelData[(panelData$GARDEN == "WG1" & panelData$TREAT == "I" & panelData$TYPE == "LN(Biomass)"), ]$cols <- "grey80"
panelData[(panelData$GARDEN == "WG1" & panelData$TYPE == "Richness" & panelData$TREAT == "I"), ]$cols <- "grey80"
panelData[(panelData$GARDEN == "WG1" & panelData$TYPE == "Richness (Woody plants)" & panelData$TREAT == "I"), ]$cols <- "grey80"

# Plot the panel
p <- ggplot(panelData, aes(x = TREAT, y = VALUE, group=GARDEN)) + 
  geom_line(aes(linetype = GARDEN), size = 0.5, alpha=0.15)+
  geom_jitter(width = 0.05, size=2, 
              shape = 19, color=panelData$cols)
p + facet_wrap(~TYPE, scales="free", drop=TRUE) +theme_bw() + 
  theme(axis.text.x=element_text(angle=0, hjust=0.5),
        strip.text = element_text(size=8),
        legend.justification=c(0.5,0.5), 
        legend.position="bottom")+xlab("")+ylab("")
```

Fig. 1 Community descriptors showing significant response to the experimental treatments for the whole community and for tree communitites (based on all tree species present in the experimental plots) in Wanang. Red colour represents significant difference at the $\alpha < 0.05$. Orange colour denotes marginal significance $\alpha < 0.1$.

```{r, results='asis'}
# Test the general significance
l1W <- lmer(log(BIO)~TREAT + (1|GARDEN), data=AllTestData)
l2W <- lmer(SW~TREAT + (1|GARDEN), data=AllTestData)
l3W <- glmer(SPEC_NO~TREAT+(1|GARDEN), data=AllTestData,
                   family="poisson")
l4W <- lmer(SIMP~TREAT + (1|GARDEN), data=AllTestData)
l5W <- lmer(EVEN~TREAT + (1|GARDEN), data=AllTestData)


# Test trees
l1Wt <- lmer(logBio~TREATMENT + (1|GARDEN), data=TreeTestDataset)
l2Wt <- lmer(DivTree~TREATMENT + (1|GARDEN), data=TreeTestDataset)
l3Wt <- glmer(species~TREATMENT+(1|GARDEN), data=TreeTestDataset,
             family="poisson")
l4Wt <- lmer(1/invSimp~TREATMENT + (1|GARDEN), data=TreeTestDataset)
l5Wt <- lmer(Evenness~TREATMENT + (1|GARDEN), data=TreeTestDataset)
l6Wt <- glmer(stems~TREATMENT + (1|GARDEN), data=TreeTestDataset,
              family="poisson")

###################################################

# Insecticide plot removed
AllTestData <- AllTestData[!(AllTestData$GARDEN == "WG1" & AllTestData$TREAT == "INSECTICIDE"), ]

# Test the general significance
l1W <- lmer(log(BIO)~TREAT + (1|GARDEN), data=AllTestData)
l2W <- lmer(SW~TREAT + (1|GARDEN), data=AllTestData)
l3W <- glmer(SPEC_NO~TREAT+(1|GARDEN), data=AllTestData,
                   family="poisson")
l4W <- lmer(SIMP~TREAT + (1|GARDEN), data=AllTestData)
l5W <- lmer(EVEN~TREAT + (1|GARDEN), data=AllTestData)

TreeTestDataset  <- TreeTestDataset [!(TreeTestDataset $GARDEN == "WG1" & TreeTestDataset $TREAT == "INSECTICIDE"), ]

# Test trees
l1Wt <- lmer(logBio~TREATMENT + (1|GARDEN), data=TreeTestDataset)
l2Wt <- lmer(DivTree~TREATMENT + (1|GARDEN), data=TreeTestDataset)
l3Wt <- glmer(species~TREATMENT+(1|GARDEN), data=TreeTestDataset,
             family="poisson")
l4Wt <- lmer(1/invSimp~TREATMENT + (1|GARDEN), data=TreeTestDataset)
l5Wt <- lmer(Evenness~TREATMENT + (1|GARDEN), data=TreeTestDataset)
l6Wt <- glmer(stems~TREATMENT + (1|GARDEN), data=TreeTestDataset,
              family="poisson")

kable(summary(l1W)$coef, digits = 3,caption = "Biomass Weevil 125")
kable(summary(l2W)$coef, digits = 3,caption = "Shannon Insecticide and marginally weevil 25")
kable(summary(l3W)$coef, digits = 3,caption = "Species number: Insecticide")
kable(summary(l4W)$coef, digits = 3,caption = "Simpsons: Insecticide")
kable(summary(l5W)$coef, digits = 3,caption = "Evenness: marginally insecticide and weevil 25")

kable(summary(l1Wt)$coef, digits = 3, caption = "Results of the linear mixed effect model of the effect of the treatment on the general productivity of the plot for trees")
kable(summary(l2Wt)$coef, digits = 3, caption = "Results of the linear mixed effect model of the effect of the treatment on the species diversity of the plot for trees")
kable(summary(l3Wt)$coef, digits = 3, caption = "Results of the linear mixed effect model of the effect of the treatment on the species richness of the plot. Errors were modeled by quasi poisson distribution because of the ovrerdispersion")
kable(summary(l4Wt)$coef, digits = 3, caption = "Results of the linear mixed effect model of the effect of the treatment on the species richness (Simpson's D) of the plot. Errors were modeled by quasi poisson distribution because of the ovrerdispersion")
kable(summary(l5Wt)$coef, digits = 3, caption = "Results of the linear mixed effect model of the effect of the treatment on the evenness of the plot")
kable(summary(l6Wt)$coef, digits = 3, caption = "Results of the linear mixed effect model of the effect of the treatment on the number of stems per plot.Errors were modeled by quasi poisson distribution because of the ovrerdispersion")

```


# Community weighted means

## The whole community
```{r, fig.height = 4, fig.width=5}

# add colours 
# herb Nothing
# ldmc Significantlyhigher for I
panelCWM[(panelCWM$Type == "LDMC" & panelCWM$Treat == "I" ), ]$cols <- sig
# sla Marginally lower for I
panelCWM[(panelCWM$Type == "SLA" & panelCWM$Treat == "I" ), ]$cols <- msig

# herbt significantly higher for P and H2
panelCWM[(panelCWM$Type == "Herbivory (Woody plants)" & panelCWM$Treat == "P" ), ]$cols <- sig
panelCWM[(panelCWM$Type == "Herbivory (Woody plants)" & panelCWM$Treat == "H2" ), ]$cols <- sig

# ldmct significantly higher I
panelCWM[(panelCWM$Type == "LDMC (Woody plants)" & panelCWM$Treat == "I" ), ]$cols <- sig

# slat No effect

# EXCLUDED INSECTICIDE PLOT
panelCWM[(panelCWM$GARDEN == "WG1" & panelCWM$Treat == "I" ), ]$cols <- "grey80"

# Exclude nonsignificant plots
panelCWMss <- panelCWM[panelCWM$Type %in% c("SLA", "LDMC",
                              "LDMC (Woody plants)",
                              "Herbivory (Woody plants)"), ]

# Plots for CWM's (Herbivory, SLA, LDMC)
cwm <- ggplot(panelCWMss, aes(x = Treat, y = Value, group=GARDEN)) + 
  geom_line(aes(linetype = GARDEN), size = 0.5, alpha=0.15)+
  geom_jitter(width = 0.05, size=2, 
              shape = 19, color=panelCWMss$cols)
cwm + facet_wrap(~Type, scales="free", drop=TRUE) +theme_bw() + 
  theme(axis.text.x=element_text(angle=0, hjust=0.5))+xlab("")+ylab("")

# Test without W1
panelTest <- panelCWM[!(panelCWM$GARDEN == "WG1" & panelCWM$Treat == "I"),]
panelTest$Treat <- relevel(panelTest$Treat, ref = "C")

herb <- lmer(Value~Treat + (1|GARDEN), data=panelTest[panelTest$Type == "Herbivory",])
ldmc <- lmer(Value~Treat + (1|GARDEN), data=panelTest[panelTest$Type == "LDMC",])
sla  <- lmer(Value~Treat + (1|GARDEN), data=panelTest[panelTest$Type == "SLA",])

herbt <- lmer(Value~Treat + (1|GARDEN), 
              data=panelTest[panelTest$Type == "Herbivory (Woody plants)",])
ldmct <- lmer(Value~Treat + (1|GARDEN), 
              data=panelTest[panelTest$Type == "LDMC (Woody plants)",])
slat  <- lmer(Value~Treat + (1|GARDEN), 
              data=panelTest[panelTest$Type == "SLA (Woody plants)",])

kable(summary(herb)$coef, caption="CWM herb, no WG1")   # Nothing
kable(summary(ldmc)$coef,caption="CWM ldmc, no WG1")   # Significantlyhigher for I
kable(summary(sla)$coef,caption="CWM sla, no WG1")    # Marginally lower for I
kable(summary(herbt)$coef, caption="CWM herb, no WG1 trees")  # significantly higher for P and H2
kable(summary(ldmct)$coef, caption="CWM ldmc, no WG1 trees")  # significantly higher I
kable(summary(slat)$coef, caption="CWM sla, no WG1 trees")   # No effect
```

Fig. 3. Community weighted mesans for three traits based on the whole community from Wanang

## Functional diversity plot.

```{r,fig.height = 4, fig.width=7}
fdDataW$GARDEN <- substr(fdDataW$plotCode,1,3)
fdDataW$cols <- "grey40"

# fdDataW$treatment <- as.character(fdDataW$treatment)
# fdDataW$fdVals.ind <- as.character(fdDataW$fdVals.ind)
# fdDataW$GARDEN <- as.character(fdDataW$GARDEN)

sig <- "red"
msig <- "orange"

# Colour for the plot
# "FRic" "FEve" "FDiv" "FDis" "RaoQ"
# kable(summary(frich_testR)$coef) # mar: F,I,P sig: W125
fdDataW[(fdDataW$fdVals.ind == "FRic" & fdDataW$treatment == "F"), ]$cols<-msig
fdDataW[(fdDataW$fdVals.ind == "FRic" & fdDataW$treatment == "I"), ]$cols<-msig
fdDataW[(fdDataW$fdVals.ind == "FRic" & fdDataW$treatment == "P"), ]$cols<-msig
fdDataW[(fdDataW$fdVals.ind == "FRic" & fdDataW$treatment == "H2"), ]$cols<-sig
# kable(summary(raoq_testR)$coef)  # mar: none  sig: F
fdDataW[(fdDataW$fdVals.ind == "RaoQ" & fdDataW$treatment == "F"), ]$cols<-sig
# kable(summary(fdis_testR)$coef)  # mar: I     sig: F
fdDataW[(fdDataW$fdVals.ind == "FDis" & fdDataW$treatment == "I"), ]$cols<-msig
fdDataW[(fdDataW$fdVals.ind == "FDis" & fdDataW$treatment == "F"), ]$cols<-sig
# kable(summary(fdiv_testR)$coef)  # mar: I     sig: none
fdDataW[(fdDataW$fdVals.ind == "FDiv" & fdDataW$treatment == "I"), ]$cols<-msig
# kable(summary(feve_testR)$coef)  # mar: none  sig: none
# wg1 insecticide grey80
fdDataW[(fdDataW$GARDEN == "WG1" & fdDataW$treatment == "I"), ]$cols <- "grey80"

fdPlot <- ggplot(fdDataW, aes(x=treatment, y=fdVals.values, group=GARDEN))+
  geom_line(aes(linetype=GARDEN), size=0.5,alpha=0.15)+
  geom_jitter(width = 0.1,size = 2.5,color=fdDataW$cols)

fdPlot + facet_wrap(~ fdVals.ind, scales="free", drop=TRUE) +theme_bw() + theme(axis.text.x=element_text(angle=0, hjust=0.5))


# Tests
# Need to validate and do all the global tests
frich_test <- lmer(FRic ~ Treat+(1|Block), data=fdDataTest)
raoq_test <- lmer(RaoQ ~ Treat+(1|Block), data=fdDataTest)
fdis_test <- lmer(FDis ~ Treat+(1|Block), data=fdDataTest)
fdiv_test <- lmer(FDiv ~ Treat+(1|Block), data=fdDataTest)
feve_test <- lmer(FEve ~ Treat+(1|Block), data=fdDataTest)

# kable(summary(frich_test)$coef) # mar: F, P. sig: I, W125
# kable(summary(raoq_test)$coef)  # mar: none  sig: F
# kable(summary(fdis_test)$coef)  # mar: I     sig: F
# kable(summary(fdiv_test)$coef)  # mar: I     sig: none
# kable(summary(feve_test)$coef)  # mar: none  sig: none

# Remove the WG1 INSECTICIDE PLOT
fdDataTestRed <- fdDataTest[!(fdDataTest$Block == "WG1" & fdDataTest$Treat == "INSECTICIDE"),]

frich_testR <- lmer(FRic ~ Treat+(1|Block), data=fdDataTestRed)
raoq_testR <- lmer(RaoQ ~ Treat+(1|Block), data=fdDataTestRed)
fdis_testR <- lmer(FDis ~ Treat+(1|Block), data=fdDataTestRed)
fdiv_testR <- lmer(FDiv ~ Treat+(1|Block), data=fdDataTestRed)
feve_testR <- lmer(FEve ~ Treat+(1|Block), data=fdDataTestRed)

kable(summary(frich_testR)$coef,
      caption="Functional richness, no WG1 I ") # mar: F,I,P sig: W125
kable(summary(raoq_testR)$coef,
      caption="Rao richness, no WG1 I ")  # mar: none  sig: F
kable(summary(fdis_testR)$coef,
      caption="Functional dispersion, no WG1 I ")  # mar: I     sig: F
kable(summary(fdiv_testR)$coef,
      caption="Functional divegence, no WG1 I ")  # mar: I     sig: none
kable(summary(feve_testR)$coef,
      caption="Functional evenness, no WG1 I ")  # mar: none  sig: none

```
Fig. 4. Functional diverstiy for all speceis and the tree communnity present in the community based on SLA, LDMC, Herbivory.

## FD characteristics based only on wood trees

```{r, fig.height = 4, fig.width=7}
fdDataTW$GARDEN <- substr(fdDataTW$plotCode,1,3)
fdDataTW$cols <- "grey40"

sig <- "red"
msig <- "orange"

# "FRic" "FEve" "FDiv" "FDis" "RaoQ"
# kable(summary(frich_testTR)$coef) # mar: F,P   sig: I
fdDataTW[(fdDataTW$fdVals.ind == "FRic" & fdDataTW$treatment == "F"),]$cols <- msig
fdDataTW[(fdDataTW$fdVals.ind == "FRic" & fdDataTW$treatment == "P"),]$cols <- msig
fdDataTW[(fdDataTW$fdVals.ind == "FRic" & fdDataTW$treatment == "I"),]$cols <- sig
# kable(summary(raoq_testTR)$coef)  # mar: none  sig: I
fdDataTW[(fdDataTW$fdVals.ind == "RaoQ" & fdDataTW$treatment == "I"),]$cols <- sig
# kable(summary(fdis_testTR)$coef)  # mar: none  sig: I
fdDataTW[(fdDataTW$fdVals.ind == "FDis" & fdDataTW$treatment == "I"),]$cols <- sig
# kable(summary(fdiv_testTR)$coef)  # mar: none  sig: none
# kable(summary(feve_testTR)$coef)  # mar: none  sig: none

# Colours for the plot NO Insecticide WG1
fdDataTW[(fdDataTW$GARDEN == "WG1" & fdDataTW$treatment == "I"),]$cols <- "grey80"

fdPlot2 <- ggplot(fdDataTW, aes(x=treatment, y=fdVals.values, group=GARDEN))+
  geom_line(aes(linetype=GARDEN), size=0.5,alpha=0.15)+
  geom_jitter(width = 0.1,size = 2.5,color=fdDataTW$cols)

fdPlot2 + facet_wrap(~ fdVals.ind, scales="free", drop=TRUE) +theme_bw() + theme(axis.text.x=element_text(angle=0, hjust=0.5))

frich_testT <- lmer(FRic ~ Treat+(1|Block), data=fdDataTwTest)
raoq_testT <- lmer(RaoQ ~ Treat+(1|Block), data=fdDataTwTest)
fdis_testT <- lmer(FDis ~ Treat+(1|Block), data=fdDataTwTest)
fdiv_testT <- lmer(FDiv ~ Treat+(1|Block), data=fdDataTwTest)
feve_testT <- lmer(FEve ~ Treat+(1|Block), data=fdDataTwTest)

kable(summary(frich_testT)$coef) # mar: W125. sig: F, I, P
kable(summary(raoq_testT)$coef)  # mar: none  sig: none
kable(summary(fdis_testT)$coef)  # mar: I     sig: none
kable(summary(fdiv_testT)$coef)  # mar: none  sig: none
kable(summary(feve_testT)$coef)  # mar: none  sig: none

# Reduced dataset
fdDataTwTestRed <- fdDataTwTest[!(fdDataTwTest$Block == "WG1" & fdDataTwTest$Treat == "INSECTICIDE"),]

frich_testTR <- lmer(FRic ~ Treat+(1|Block), data=fdDataTwTestRed)
raoq_testTR <- lmer(RaoQ ~ Treat+(1|Block), data=fdDataTwTestRed)
fdis_testTR <- lmer(FDis ~ Treat+(1|Block), data=fdDataTwTestRed)
fdiv_testTR <- lmer(FDiv ~ Treat+(1|Block), data=fdDataTwTestRed)
feve_testTR <- lmer(FEve ~ Treat+(1|Block), data=fdDataTwTestRed)

kable(summary(frich_testTR)$coef) # mar: F,P   sig: I
kable(summary(raoq_testTR)$coef)  # mar: none  sig: I
kable(summary(fdis_testTR)$coef)  # mar: none  sig: I
kable(summary(fdiv_testTR)$coef)  # mar: none  sig: none
kable(summary(feve_testTR)$coef)  # mar: none  sig: none

```

## Randomizations

```{r, fig.height = 4, fig.width=6}
dataRC_facets_plot$sig <- rep(c("NS","*","*","NS","NS"), each=2)

pl <- ggplot() +
    geom_errorbar(data=dataRC_facets_plot,
                  mapping=aes(x=treat, ymin=lose, ymax=upse), width=0.2,
                  size=1, color="gray60") + ylab("Within treatment dissimilarity (RC)") +
  xlab("")+
    geom_hline(yintercept = c(0), lty=4, lwd = 0.8, col="grey60") +
    geom_point(data=dataRC_facets_plot, mapping=aes(x=treat, y=meanRC), size=1, shape=21,
               fill="black")+
    ylim(-1,1) + geom_hline(yintercept = c(1, -1), lty=2, col="grey60") +
    facet_wrap(~type, scales="free", drop=TRUE) + theme_bw() +
    theme(axis.text.x=element_text(angle=0, hjust=0.5),
          strip.text = element_text(size=8))
pl + annotate("text", label = c("","*","*","",""), size = 10, x = 1.5, y = 0.75)
```